Assessing primitives performance\\ on multi-stage execution

### 2018-05-13 paper.tex
final pass

\paragraph{Pure Slang with JIT support.}

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if there is no direct link between readability and mere lines of code. \sd{well this is kind of weak. but I do not know what do propose. Cyclomatic complexity counts the number of different exits. And it would be a nice complement}
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead. \cba{InterpreterProxy not defined before}
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

One or another implementation could be preferred depending on the intended aim: the main trade-off will be found between the engineering time and the performance. Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time may tend to increase when more low-level languages are used, such as Cog's RTL for instance.
Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run. Likewise, the use of plugins may bring modularity, but as you will see in Section \ref{sec:eval}, the performance, although better than baseline version, could still be improved.

@@

The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM. In this section, we will compare the performances of these primitives against the other existing implementations performances. We have measures their respective execution times for different string sizes.

@@

   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
   \hline
  \multirow{2}{*}{SmartSyntax} &GCC & 0.3411 +- 0.0007 & 0.6014 +- 0.0023 & 1.3891 +- 0.0023 &  8.6796 +- 0.0587 & 13.6051 +- 0.0589\\
 							   &LLVM & 0.3147 +- 0.0023 & 0.5606 +- 0.0022 & 1.2505 +- 0.0046 & 10.3029 +- 0.0624 & 9.4065 +- 0.0367\\
   \hline
  \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
  							   &LLVM& 0.3036 +- 0.0018 & 0.5567 +- 0.0010 & 1.2296 +- 0.0018 & 6.6033 +- 0.0248  & 9.2138 +- 0.0140 \\
   \hline
  \multirow{2}{*}{PureSlang} &GCC& 1.0778 +- 0.0054 & 1.8169 +- 0.0191 & 3.8190 +- 0.0545 & 15.8095 +- 0.3005 & 24.4178 +- 0.1863\\
 							 &LLVM& 0.8545 +- 0.0061 & 1.5605 +- 0.0064 & 2.8205 +- 0.0080 & 10.3029 +- 0.0615  & 13.4385 +- 0.0402 \\
   \hline
  \multirow{2}{*}{Slang+JIT} &GCC& 2.4904 +- 0.0043 & 3.5378 +- 0.0124  & 6.7405 +- 0.0108 & 16.7527 +- 0.1063 & 24.4801 +- 0.0950\\
 							&LLVM & 2.2507 +- 0.0413 & 3.7545 +- 0.0111 & 7.0898 +- 0.0140 & 17.9796 +- 0.1013  & 26.4048 +- 0.0930 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline}
\label{tab:detail_bench}
\end{table*}

@@

The two compared strings are equal as it corresponds to the worst scenario for most of our benched implementations. There is one less branch in the two plugin implementations in this case.

@@

%\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand. Figure \ref{fig:bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for GCC and LLVM.

We can draw the following patterns from these results:
\begin{itemize}
\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than the baseline's for the small strings.
\item The Slang+JIT implementation is more efficient than the baseline in every case. Nevertheless,the performance gap between this version and the PureSlang one tends to decrease as the string size increases. 
\item The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings. 
\end{itemize}

We can explain these results based on the last 3 criteria detailed in Section \ref{sec:implem_cmp}:

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure \ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performances are close to the baseline's.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time.

\paragraph{Optimization potential on large strings comparison.}
The plugins come with an overhead \sk{not an overhead when it comes to inlining. Just not optimized. need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section~\ref{sec:implem_cmp}. This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure \ref{fig:Asm}). However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster while Slang+JIT is 16.8 times faster. This is not the case anymore on 1000-characters strings.
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string. Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.

This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure \ref{fig:Asm}, GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors. On the contrary, LLVM generates more instructions: it especially fails at preserving the registries holding the strings (edi and esi), leading to 2 extra instructions (bolded in the figure) that will load the strings at each iteration.

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitive is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the performance difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the performance difference between all three implementation was not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.

These approaches have the advantage of decreasing the primitive maintenance cost since fewer primitives have to be maintained. The performance problem is however partially solved: only the peak performance is similar with and without the primitives. Since those system rely on JITs, start-up performance is worse than peak performance, increasing the performance difference at start-up between the client code and the primitive. In addition, the performance can be  unreliable with optimizing JITs. The implementors of Morphic \cite{MorphicSelf} on top of the Self VM were complaining that changing a single line of code could lead to important performance drop and they would not understand why. Those performance drops were due to the optimizing JIT taking different optimization decisions. This problem was so significant that recently the V8 team \cite{V8} rewrote their primitives written in Javascript (named built-ins in their project) in a DSL compiled ahead of time through their JIT back-end to improve the performance of the primitives and therefore decrease the performance drop between baseline and peak performance. Aside from the start-up performance and the performance unreliability problems described, these approaches also require a mature optimizing JIT, hard to implement and maintain.

@@

To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition in a DSL compiled ahead of time through their JIT back-end to native code. With their primitives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives). In this context, the stack switch overhead is not a problem anymore.

@@

All the primitive we compare are written in a processor independent way, due to the C compiler abstraction for the Slang versions, the abstract assembly abstraction for Cog's RTL version or the VM abstraction for the baseline implementation in Smalltalk. We could rewrite the primitive differently for each back-end currently supported (x86, x64, MIPSEL, ARMv6) and yet to support (ARMv8, ...). It could be relevant in some cases performance-wise, for example, on intel processors, the JVM implements string comparison using the SSE4.2 string comparison instructions when available \cite{JVMSSE4StringComp}. We did not go in this direction because the maintenance cost in this case is definitely too high since we would have to maintain a different implementation per back-end  (already 4 implementations today).

>>>>

\paragraph{Pure Slang with JIT support.}\sd{Why not calling it Slang+RTL}

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if there is no direct link between readability and mere lines of code. \sd{well this is kind of weak. but I do not know what do propose. Cyclomatic complexity counts the number of different exits. And it would be a nice complement}
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead. \cba{InterpreterProxy not defined before}
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

One or another implementation could be preferred depending on the intended aim: the main trade-off is between  engineering time and  performance. Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used, such as Cog's RTL for instance.
Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run. Likewise, the use of plugins brings modularity, but as you will see in Section~\ref{sec:eval}, the performance, although better than baseline version, could still be improved.

@@

The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes. \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
   \hline
  \multirow{2}{*}{SmartSyntax} &GCC & 0.3411 +- 0.0007 & 0.6014 +- 0.0023 & 1.3891 +- 0.0023 &  8.6796 +- 0.0587 & 13.6051 +- 0.0589\\
 							   &LLVM & 0.3147 +- 0.0023 & 0.5606 +- 0.0022 & 1.2505 +- 0.0046 & 10.3029 +- 0.0624 & 9.4065 +- 0.0367\\
   \hline
  \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
  							   &LLVM& 0.3036 +- 0.0018 & 0.5567 +- 0.0010 & 1.2296 +- 0.0018 & 6.6033 +- 0.0248  & 9.2138 +- 0.0140 \\
   \hline
  \multirow{2}{*}{PureSlang} &GCC& 1.0778 +- 0.0054 & 1.8169 +- 0.0191 & 3.8190 +- 0.0545 & 15.8095 +- 0.3005 & 24.4178 +- 0.1863\\
 							 &LLVM& 0.8545 +- 0.0061 & 1.5605 +- 0.0064 & 2.8205 +- 0.0080 & 10.3029 +- 0.0615  & 13.4385 +- 0.0402 \\
   \hline
  \multirow{2}{*}{Slang+JIT} &GCC& 2.4904 +- 0.0043 & 3.5378 +- 0.0124  & 6.7405 +- 0.0108 & 16.7527 +- 0.1063 & 24.4801 +- 0.0950\\
 							&LLVM & 2.2507 +- 0.0413 & 3.7545 +- 0.0111 & 7.0898 +- 0.0140 & 17.9796 +- 0.1013  & 26.4048 +- 0.0930 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline}
\label{tab:detail_bench}
\end{table*}

@@

The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. There is one less branch in the two plugin implementations in this case.

@@

%\cba{Je pense string de taille 0-1 c'est cool pour montrer que le code C va moins vite que pur smalltalk dans ce cas (si c'est vraiment le cas)}
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters, using two different compilers: GCC on one hand and LLVM on the other hand. Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for GCC and LLVM.

We can draw the following patterns from these results:
\begin{itemize}
\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than \sd{within which ranges? 10\% 200\%} the baseline's for the small strings.
\item The Slang+JIT implementation is more efficient than the baseline in every case. Nevertheless, the performance gap between this version and the PureSlang one tends to decrease as the string size increases. 
\item The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings. 
\end{itemize}

We can explain these results based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}:

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time \sd{Why that? just during the time the JIT does not kick in?}.

\paragraph{Optimization potential on large strings comparison.}
The plugins come with an overhead \sk{not an overhead when it comes to inlining. Just not optimized. need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section~\ref{sec:implem_cmp}. This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}). However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster \sd{than what} while Slang+JIT is 16.8 times faster \sd{than what}. This is not the case anymore on 1000-characters strings.
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performance between the PureSlang and Slang+JIT implementations is really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string. Yet performance greatly differs when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.

This difference is mainly due to the optimizations performed by both the compilers: as depicted in Figure~\ref{fig:Asm}, GCC generates instructions that are very close to the ones generated by the JIT, hence the similar speed-up factors. On the contrary, LLVM generates more instructions: it especially fails at preserving the registries \sd{registers?} holding the strings (edi and esi), leading to two extra instructions (bolded in the figure) that  load the strings at each iteration.

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitive is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.

These approaches have the advantage of decreasing the primitive maintenance cost since fewer primitives have to be maintained. The performance problem is however partially solved: only the peak performance is similar with and without the primitives. Since those systems rely on JITs, start-up performance is worse than peak performance, increasing the performance difference at start-up between the client code and the primitive. In addition, the performance can be  unreliable with optimizing JITs. The implementors of Morphic \cite{MorphicSelf} on top of the Self VM were complaining that changing a single line of code could lead to important performance drop and they would not understand why. Those performance drops were due to the optimizing JIT taking different optimization decisions. This problem was so significant that recently the V8 team \cite{V8} rewrote their primitives written in Javascript (named built-ins in their project) in a DSL compiled ahead of time through their JIT back-end to improve the performance of the primitives and therefore decrease the performance drop between baseline and peak performance. Aside from the start-up performance and the performance unreliability problems described, these approaches also require a mature optimizing JIT, hard to implement and maintain.

@@

To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition \sd{what it is?} in a DSL compiled ahead of time through their JIT back-end to native code. With their primitives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives). In this context, the stack switch overhead is not a problem anymore.

@@

All the primitives we compared are written in a processor independent way, due to the C compiler abstraction for the Slang versions, the abstract assembly abstraction for Cog's RTL version or the VM abstraction for the baseline implementation in Smalltalk. We could rewrite the primitive differently for each back-end currently supported (x86, x64, MIPSEL, ARMv6) and yet to support (ARMv8, ...). It could be relevant in some cases performance-wise, for example, on Intel processors, the JVM implements string comparison using the SSE4.2 string comparison instructions when available~\cite{JVMSSE4StringComp}. We did not go in this direction because the maintenance cost in this case is definitely too high since we would have to maintain a different implementation per back-end (already 4 implementations today \sd{that we have or that we would have}).
